---
title: "Getting started with ACTman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ACTman}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette walks through a complete ACTman analysis using a real,
anonymized actigraphy recording bundled with the package, rather than
synthetic data. If you use ACTman in published work, please cite the paper
describing it:

> Kunkels, Y. K., Knapen, S. E., Zuidersma, M., Wichers, M., Riese, H., &
> Emerencia, A. C. (2020). ACTman: Automated preprocessing and analysis of
> actigraphy data. *Journal of Science and Medicine in Sport*, 23(5),
> 481-486. <https://doi.org/10.1016/j.jsams.2019.11.009>

## The example data

ACTman bundles a real MotionWatch 8 (CamNtech) recording, in two forms:

- `system.file("extdata", "example-mw8-participant.csv", package =
  "ACTman")`: a ~7-day subset in genuine raw MW8 export format (device
  metadata lines, a `Raw data:` marker, then `Date,Time,Activity` rows at
  30-second epochs) -- exactly what you'd get from a real MotionWare
  export, useful for seeing the full pipeline including file reading.
- `data(ACTdata.1)`: the complete ~33-day recording (2017-07-05 to
  2017-08-07, 94,721 observations), already extracted to the
  Date/Time/Activity form ACTman produces internally partway through its
  pipeline.

This vignette uses the raw-format file, since that's what you'd actually
point `ACTman()` at with your own data.

```{r setup}
library(ACTman)

example_file <- system.file("extdata", "example-mw8-participant.csv", package = "ACTman")
workdir <- tempfile("actman-vignette-")
dir.create(workdir)
file.copy(example_file, file.path(workdir, "participant01.csv"))

list.files(workdir)
```

## Running the analysis

A single `ACTman()` call reads the file, bins any 30-second epochs into
60-second epochs, checks for missing data and trailing zero-activity, and
computes the non-parametric circadian rhythm variables:

```{r run-actman}
result <- ACTman(
  workdir = workdir,
  myACTdevice = "MW8",
  circadian_analysis = TRUE,
  iwantsleepanalysis = FALSE,
  plotactogram = FALSE,
  lengthcheck = FALSE # this example file is only ~7 days; don't require 14
)

result
```

`ACTman()` always returns a single `actman_result` object; here, since we
asked for a circadian analysis only, `$circadian` is populated and
`$sleep`/`$rolling_window` are `NULL`.

## Interpreting the circadian rhythm variables

```{r circadian-vars}
result$overview[, c("filename", "recordingtime2", "missings_perc")]
result$circadian
```

- **IS** (Interdaily Stability) close to 1 means a highly repeatable
  24-hour activity pattern day after day; closer to 0 means little
  consistent daily structure.
- **IV** (Interdaily Variability) captures fragmentation: higher values
  mean more switching between rest and activity within a day.
- **L5**/**M10** are the average activity levels during the 5 least-active
  and 10 most-active hours of the average day, with their respective
  onset times (`L5_starttime`, `M10_starttime`).
- **RA** (Relative Amplitude), `(M10 - L5) / (M10 + L5)`, summarizes the
  day/night activity contrast in a single normalized number.

See `vignette("actman-intro")`'s references, or `?circadian_metrics`, for
the methodological background (Van Someren et al., 1999; Witting et al.,
1990).

## Actogram

```{r actogram, fig.alt = "24-hour actogram of the example participant, showing one horizontal activity bar per day with a clear day/night rest-activity pattern"}
ACTman(
  workdir = workdir,
  myACTdevice = "MW8",
  circadian_analysis = FALSE,
  iwantsleepanalysis = FALSE,
  plotactogram = "24h",
  lengthcheck = FALSE
)

actogram_file <- list.files(file.path(workdir, "Actograms"), full.names = TRUE, pattern = "\\.pdf$")[1]
actogram_file
```

`plot_actogram()` writes a PDF (24h or 48h) to `workdir/Actograms`; open
`r basename(actogram_file)` to see it (PDFs can't be embedded inline in
this vignette, but see the actogram image in the package's `README.md` for
an example of what it looks like).

## Moving-window analysis

For longer recordings, `movingwindow = TRUE` re-runs the circadian (and
early-warning-signal) calculations on a sliding window instead of the
whole recording at once, producing a time series of each metric. This
needs more days of data than the 7-day example file provides for a
window of the default size, so this example uses a 3-day window stepped
by 1 day:

```{r moving-window}
rolling_result <- ACTman(
  workdir = workdir,
  myACTdevice = "MW8",
  circadian_analysis = FALSE,
  iwantsleepanalysis = FALSE,
  plotactogram = FALSE,
  movingwindow = TRUE,
  movingwindow.size = 3,
  movingwindow.jump = 1,
  lengthcheck = FALSE
)

rolling_result$rolling_window[, c("starttime", "endtime", "IS", "IV", "RA", "Mean", "SD")]
```

Each row is one 3-day window; `IS`/`IV`/`RA` are the same circadian
variables as before, now tracked across the recording, alongside
early-warning-signal statistics (`Mean`, `SD`, autocorrelation, etc. --
see `?ews_metrics`) that have been explored as potential indicators of
mood-state transitions in longer recordings (Van de Leemput et al., 2014).

## Sleep analysis

Sleep analysis (`iwantsleepanalysis = TRUE`) additionally requires a
sleeplog or event-marker file alongside the actigraphy data (see the "Data
formats" section of `README.md` for the expected format) -- not included
with this particular example recording, so it isn't demonstrated here. See
`?sleepdata_overview` and `?sleeplog_from_markers` for details.

## Cleaning up

```{r cleanup, include = FALSE}
unlink(workdir, recursive = TRUE)
```
